---
title: "*Brazilian 2018 Elections EDA*"
# author: "Gaston Guillaux"
# date: "2020-01-26"
output: html_notebook
---

In this notebook we will cover an approach to do an exploratory data analysis with the data of the Brazilian general elections of 2018.

## **1. Obtaining the data**
We will fetch the data from the oficial governemnt repository available at the webpage of the TSE (Tribunal Superior Eleitoral.)

* **Main URL** - http://www.tse.jus.br/eleicoes/estatisticas/repositorio-de-dados-eleitorais-1/repositorio-de-dados-eleitorais
* **Main frame from above URL** - http://www.tse.jus.br/hotsites/pesquisas-eleitorais/index.html
* **Tab with the accountability data** - http://www.tse.jus.br/hotsites/pesquisas-eleitorais/prestacao_contas.html
* **Final link with the 2018 data** - http://www.tse.jus.br/hotsites/pesquisas-eleitorais/prestacao_contas_anos/2018.html

### 1.1 Importing libraries
```{r}
require(rvest)     # web scrapping
require(plotly)    # data visualization
require(ggplot2)   # data visualization
require(lubridate) # format dates 
require(tidyverse) # data cleansing
```

### 1.2 Set working directory {.tabset}
#### **a) Variables**
```{r}
wd       <- '~/dev/r' # '/cloud/project'
project  <- 'brz_elections'
raw_data <- 'raw_data'
```

#### **b) First level path**
```{r}
dest_path <- file.path(wd, project)

if (!dir.exists(dest_path)) {
  dir.create(dest_path)   # create path case it does not exist
  print(paste("Folder", dest_path , "created at", wd))
} else {
  print(paste("Folder", dest_path , "already exists"))
}
```

#### **c) Second level path**
```{r}
dest_path <- file.path(dest_path, raw_data)

if (!dir.exists(dest_path)) {
  dir.create(dest_path) # create directory
  print(paste("Folder ", dest_path , "created sucessfully "))
} else {
  print(paste("Folder ", dest_path , "already exists"))
}
```

#### **d) Set working directory**
```{r}
setwd(dest_path)
```

### 1.3 Dowload source files {.tabset}
#### **a) Variables**
```{r}
# main page with the source data
target_url <- "http://www.tse.jus.br/hotsites/pesquisas-eleitorais/prestacao_contas_anos/2018.html"

# get html from source URL above
html       <-  read_html(target_url)

# use css selector to just filter the path with the other url adresses
nodes      <-  html_nodes(html, 'div p a')

# final fetch of the url addresses with the raw data
zip_url    <-  html_attr(nodes, 'href')
```

#### **b) Download files**
```{r}
# now need to do a for loop in this case
# r handles the download of the whole vector
# *** WARNING ***
# some files are huge - >= 200 MB
download.file(url = zip_url, 
              destfile = file.path(dest_path, basename(zip_url)),
              mode = 'wb')
```

#### **c) Unziping**
```{r}

for (f in file.path(dest_path, basename(zip_url))) {
  print(f)
  unzip(f, exdir = dest_path)
}

```

#### **d) Delete some files**
```{r}
# delete redundant files that will not be used
pattern         <- '*BRASIL*'
files_to_remove <- grep(dir(dest_path), pattern = pattern, inv=T, value = T)
file.remove(file.path(dest_path,files_to_remove))
```


## **2. Loading and cleaning the data**

### 2.1 Load and clean downloaded csvs {.tabset}
#### **a) Revenues**
```{r}
# read revenues csv
receita <-  read.csv(file.path(dest_path, 'receitas_candidatos_2018_BRASIL.csv'),
                    sep = ';', 
                    dec = ',',
                    encoding = 'latin1')

# we have 57 columns! 
# to keep things simple let's remove the ones we know are not useful right now
receita$SG_UF                       <- NULL
receita$NR_CPF_VICE_CANDIDATO       <- NULL
receita$CD_ELEICAO                  <- NULL
receita$DS_ELEICAO                  <- NULL
receita$CD_TIPO_ELEICAO             <- NULL
receita$NM_TIPO_ELEICAO             <- NULL
receita$ANO_ELEICAO                 <- NULL
receita$DT_GERACAO                  <- NULL
receita$HH_GERACAO                  <- NULL
receita$TP_PRESTACAO_CONTAS         <- NULL
receita$DS_CARGO_CANDIDATO_DOADOR   <- NULL
receita$CD_ESFERA_PARTIDARIA_DOADOR <- NULL
receita$DS_ESFERA_PARTIDARIA_DOADOR <- NULL
receita$NM_DOADOR                   <- NULL
receita$CD_MUNICIPIO_DOADOR         <- NULL
receita$NM_MUNICIPIO_DOADOR         <- NULL        
receita$SQ_CANDIDATO_DOADOR         <- NULL
receita$NR_CANDIDATO_DOADOR         <- NULL
receita$NR_DOCUMENTO_DOACAO         <- NULL
receita$SQ_PRESTADOR_CONTAS         <- NULL
receita$SQ_RECEITA                  <- NULL
receita$CD_CARGO_CANDIDATO_DOADOR   <- NULL
receita$NR_CPF_CANDIDATO            <- NULL
receita$NR_CPF_CNPJ_DOADOR          <- NULL
receita$NM_PARTIDO                  <- NULL
receita$NR_PARTIDO_DOADOR           <- NULL
receita$NM_PARTIDO_DOADOR           <- NULL
receita$NR_PARTIDO                  <- NULL
receita$CD_FONTE_RECEITA            <- NULL
receita$CD_ORIGEM_RECEITA           <- NULL
receita$CD_NATUREZA_RECEITA         <- NULL
receita$CD_ESPECIE_RECEITA          <- NULL
receita$CD_CNAE_DOADOR              <- NULL
receita$SG_PARTIDO_DOADOR           <- NULL

# format dates using lubridate
receita$DT_ELEICAO            <- dmy(receita$DT_ELEICAO)
receita$DT_PRESTACAO_CONTAS   <- dmy(receita$DT_PRESTACAO_CONTAS)
receita$DT_RECEITA            <- dmy(receita$DT_RECEITA)

# display head
head(receita)
```

#### **b) Expenses paid**
```{r}
# another way to do the cleansing in r, now using expenses data
despesa <-  read.csv(file.path(dest_path, 'despesas_pagas_candidatos_2018_BRASIL.csv'),
                    sep = ';', 
                    dec = ',',
                    encoding = 'latin1')

rm_columns <- c('NR_CPF_VICE_CANDIDATO','SG_UF','CD_ELEICAO','DS_ELEICAO',
                'CD_TIPO_ELEICAO','NM_TIPO_ELEICAO','ANO_ELEICAO','DT_GERACAO',
                'HH_GERACAO','TP_PRESTACAO_CONTAS','SQ_PRESTADOR_CONTAS','NR_CPF_CANDIDATO',
                'NM_PARTIDO','NR_PARTIDO','DS_CNAE_FORNECEDOR','CD_TIPO_FORNECEDOR','CD_CNAE_FORNECEDOR',
                'DS_ESFERA_PART_FORNECEDOR','CD_ESFERA_PART_FORNECEDOR','SG_UF_FORNECEDOR',
                'CD_MUNICIPIO_FORNECEDOR','NM_MUNICIPIO_FORNECEDOR','SQ_CANDIDATO_FORNECEDOR',
                'NR_CANDIDATO_FORNECEDOR','CD_CARGO_FORNECEDOR','DS_ORIGEM_DESPESA',
                'CD_ORIGEM_DESPESA','SQ_PARCELAMENTO_DESPESA')

despesa[, rm_columns] <-NULL
head(despesa)
```

#### **c) Expenses hired**
```{r}
despesa_c <-  read.csv(file.path(dest_path, 'despesas_contratadas_candidatos_2018_BRASIL.csv'),
                    sep = ';', 
                    dec = ',',
                    encoding = 'latin1')

rm_columns <- c('NR_CPF_VICE_CANDIDATO','SG_UF','CD_ELEICAO','DS_ELEICAO',
                'CD_TIPO_ELEICAO','NM_TIPO_ELEICAO','ANO_ELEICAO','DT_GERACAO',
                'HH_GERACAO','TP_PRESTACAO_CONTAS','SQ_PRESTADOR_CONTAS','NR_CPF_CANDIDATO',
                'NM_PARTIDO','NR_PARTIDO','DS_CNAE_FORNECEDOR','CD_TIPO_FORNECEDOR','CD_CNAE_FORNECEDOR',
                'DS_ESFERA_PART_FORNECEDOR','CD_ESFERA_PART_FORNECEDOR','SG_UF_FORNECEDOR',
                'CD_MUNICIPIO_FORNECEDOR','NM_MUNICIPIO_FORNECEDOR','SQ_CANDIDATO_FORNECEDOR',
                'NR_CANDIDATO_FORNECEDOR','CD_CARGO_FORNECEDOR','DS_ORIGEM_DESPESA',
                'CD_ORIGEM_DESPESA','SQ_PARCELAMENTO_DESPESA')


despesa_c[, rm_columns] <-NULL
head(despesa_c)
```

#### **d) Backup clean frames**
```{r}
saveRDS(receita,   file.path(dest_path, 'clean_revenue.rds'))
saveRDS(despesa,   file.path(dest_path, 'clean_expenses_paid.rds'))
saveRDS(despesa_c, file.path(dest_path, 'clean_expenses_hired.rds'))
```


### 2.2 Download candidate's assets data {.tabset}
#### **a) Download**
```{r}
# download candidate asset data
cand_url <- "http://www.tse.jus.br/hotsites/pesquisas-eleitorais/candidatos_anos/2018.html"

# get html page
html       <-  read_html(cand_url)

# use selector to filter nodes
nodes      <-  html_nodes(html, 'div p a')

# get zip names
zip_url    <-  html_attr(nodes, 'href')

# download zip files
download.file(url = zip_url, 
              destfile = file.path(dest_path, basename(zip_url)),
              mode = 'wb')
# unzip
for (f in file.path(dest_path, basename(zip_url))) {
  print(f)
  unzip(f, exdir = dest_path)
}

```

#### **b) Load and clean**
```{r}
# read csvs
bem    <- read.csv(file.path(dest_path,'bem_candidato_2018_BRASIL.csv'),
                   sep = ';', 
                   dec = ',',
                   encoding = 'latin1')

status <- read.csv(file.path(dest_path,'consulta_cand_2018_BRASIL.csv'),
                   sep = ';', 
                   dec = ',',
                   encoding = 'latin1')

# clean asset frames using dplyr
c_to_remove      <- c('DT_GERACAO', 'HH_GERACAO', 'ANO_ELEICAO', 'CD_ELEICAO', 'DT_ELEICAO', 'SG_UF', 'CD_TIPO_ELEICAO',
                      'NM_TIPO_ELEICAO', 'NR_ORDEM_CANDIDATO')
bens_clean <- bem %>% select(-c_to_remove) %>%                   # seleciona todas as colunas exceto a do vetor acima
                  mutate_at(vars(starts_with('DT_')), funs(dmy)) # formata todas as datas no dataframe com lubridate


c_to_remove         <- c('DT_GERACAO', 'HH_GERACAO', 'ANO_ELEICAO', 'CD_ELEICAO', 'CD_TIPO_ELEICAO', 'NM_SOCIAL_CANDIDATO',
                         'CD_MUNICIPIO_NASCIMENTO', 'CD_ESTADO_CIVIL', 'CD_GRAU_INSTRUCAO', 'CD_COR_RACA', 'SG_UF',
                         'NR_PROTOCOLO_CANDIDATURA', 'NR_PROCESSO', 'CD_SIT_TOT_TURNO', 'CD_CARGO', 'CD_SITUACAO_CANDIDATURA',
                         'CD_GENERO', 'CD_OCUPACAO', 'NR_DESPESA_MAX_CAMPANHA')

# clean status frame
status_clean  <- status %>% select(-c_to_remove) %>% 
                            mutate_at(vars(starts_with('DT_')), funs(dmy))

```

#### **c) Backup**
```{r}
saveRDS(bens_clean, file.path(dest_path, 'clean_cand_assets.rds'))
saveRDS(status_clean, file.path(dest_path, 'clean_cand_status.rds'))

```